A generalization of the Bradley-Terry model for draws in chess

### 2019-05-15 inst/collusion_chess_draws.Rnw
work

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to chess-specific considerations in a number of ways.  Here I present
a new generalization of Bradley-Terry in which a chess match is
regarded as a three-way competition between the two players and an
entity that wins if the match is drawn.  Bradley-Terry can be further
modified to account for the white player advantage in a number of
ways.  Such reification affords insight into players' strengths,
response to playing black or white, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude that the
scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
evidence of collusion, in agreement with previous work.  However, I
find strong evidence of increased Soviet proclivity to draw in the
Interzonal Tournament, Stockholm 1962.
}

@@

\SweaveOpts{}
\begin{document}
\section{Introduction}

@@

Bradley-Terry is not directly applicable to Chess for two reasons:
firstly, the white player is at a considerable
advantage~\cite{asdfasdffdsfds}; and secondly, many chess matches end
in a draw.  Considering white advantage first, we find an overall mean
of about $(+0.375=0.394-0.276)$.  However, interpretations of this
statistic vary and the issue is hotly debated with opinions ranging
from Rowson who compares first-move advantage to that of serving in
tennis, to that of commentators such as \citeauthor{adorjan} who
claims that white's advantage is a ``delusion''; and that Black can
mount an intelligent and dynamic response to white: ``chess is a game
of complete information, and Black's information is always
greater---by one move!''.

Chess games may be won by either player, or may be drawn. The
implication of a draw is that neither player is better than the other.
Before 1867, tournament games that were drawn were replayed and in
this context we note that this procedure reduces a chess match to a
Bernoulli trial, with probability one.  This system became unworkable
due to the large number of draws, resulting in the modern practice of
awarding each player of a drawn match half a point, rendering chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are usually poorer in spectacular or striking elements such as
sharp tactical combinations or sacrifice.  Excessive drawing dissuades
sponsors, leaves spectators feeling cheated, and gives commentators
less material to discuss.  Players themselves do not seem to value
draws in the wider context of the game: \citeauthor{fischer1969}, for
example, includes only 8 draws in his compilation of 60 memorable
games, surely not representative of tournament play in which about
34\% of games are drawn ($\chi^2$, $p<3\times 10^{-6}$).

\cite{moul2009} conceptualise a player as being able to adopt a
strategy on a spectrum ranging from conservative to aggressive, thus
being able to exert some control over the probability of a draw.  They
suggest that risk-averse players may pursue less vigorous strategies,
trading expected payout for reduced variance. \cite{langer2013} {\bf
  cite something better here} point out that rational players in an
ongoing competitive situation modify their strategies depending on
whether they are leading or lagging: the lagging player will adopt
higher risk strategy (`nothing to lose'), while the leading player is
incentivised to protect the lead by lowering the risk of losing.  In a
chess context, this means the leading player will aim for a draw.

Such arguments suggest that considering a drawn chess match as a
qualitatively different outcome from a win for either player: the play
itself tends to be different, the drawing players have a different
mindset, and drawn games have less value in the wider context of the
game.

\section{Bradley-Terry}

@@

\begin{equation}
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j}\right)^{w_{ij}}
\left(\frac{p_j}{p_i+p_j}\right)^{l_{ij}}
\end{equation}

where $w_{ij}$ is the number of times $i$ beats $j$ and $l_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess match is regarded as
a three-way competition between the two players and an entity that
wins if the match is drawn.  The system is further modified to account
for first-move advantage by positing an entity whose strength is added
to that of the white player.  This methodology affords insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.

If we consider a chess match to be a three-way trial, we define a
likelihood function for non-negative $p_1,\ldots,p_n,D$ with
$D+\sum p_i=1$ and this would be

@@

\subsection{Discussion}

The above likelihood functions are objective and reasonable
representations of chess match results.  However, they differ markedly
from previous likelihood functions of draws in that they include
entities representing proclivity to draw $D$, and first-mover
advantage ($W$).

Considering white player a

The method of \citet{Edwards1972} involves only
numerical optimization (not integration which is computationally
expensive), and because gradients are available convergence is
generally rapid.

@@

Grandmasters Anand, Karpov, Kasparov have long-standing rivalry and
have played many matches amongst themselves.  A summary may be found
in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &   5 &   11  \\
  Kr & 13 &  -  &    9  \\
  Ks &  2 &   7 &  -   \\
\end{tabular}
\caption{Results of 365 chess matches \label{kka} between Grandmaster
  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  Thus, for
  example, Anand played white against Karpov a total of $18+20+13=51$
  times and won 18, drew 20, lost 13}
\end{table}

@@

and this is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
case: the likelihood for this hypothesis would be identically zero,
owing to the nonzero term in the numerator of the last term of
Equation~\ref{likelihood_WD}.  However, we see that the maximum
likelihood estimate for Draw strength $\hat{D}\simeq 0.53$: compare
the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 matches.  The MLE is lower because the
maximization process automatically adjusts for the different players'
strengths.  Secondly, we may make inferences about $W$, the extra
strength conferred by playing White: it is conceivable that $W=0$ as
this term does not appear on its own in the numerator of any term.  We
may test $H_0\colon W=0$ using the method of support by maximizing the
likelihood subject the constraint $W=0$ and comparing with the
likelihood when the optimization is free.

@@

The result of numerical optimization is 7.304, that is, well over the
two-units-of-support per degree of freedom suggested by
\citeauthor{edwards1972}; we can convincingly reject the hypothesis that
$W=0$ and infer that playing white confers a significant advantage.
This is consistent with expectation, but it is now possible to go
significantly beyond mere existence of the effect and to quantify it,
accounting for the differential strengths of the players.  Taking the
MLE above, and considering a match between say Karpov and Anand, the
strengths are Karpov=0.064, Anand=0.151, white =0.104 and draw =0.53.
Thus we would estimate that Karpov playing white against Kasparov
would win with probability
$\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
probability $\simeq 0.174$ (compare $+0.075\,=0.627-0.297$ if Karpov
plays black).

@@

One very natural and interesting hypothesis would be that the three
players are of equal strength
%equal_strength_is_assessed_in_chunk_kka_equal_strength_above____________
and we might hypothesise that
$p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  This too is readily
testable by the device of maximizing the support, subject to the
constraint that $p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  In this
case the extra support in going from the constrained to the free
optimization is about 4.93, or a likelihood ratio of about $138$.
However, this time the constraint corresponds to a loss of {\em two}
degrees of freedom so we would need 4 units of support, which is
attained in this case.  Alternatively we might follow Wilks, and
observe that the value of $2\mathcal{S}=2*4.92=9.84$ lies in the tail
region of its asymptotic distribution with a p-value of about 0.0072.
We can justifiably assert that the players do indeed have different
strengths, even if we account for white advantage and the proclivity
to draw.

@@

Players with a cautious style tend to draw CITATION NEEDED and this is
an important and interesting component of the game.  To investigate
this we need a likelihood function that can account for possible
differences in draw proclivity among the players.  This may be
achieved by considering a match between two players to be a contest
between five entities: the two real players, the white advantage, and
two personalised draw entities, one per real player; a drawn match is
counted as a win for either of the draw ghosts.  Under these
assumptions, table~\ref{kka} translates into the following likelihood
function:

@@

In the likelihood function above we see, for example, that Anand won a
total of $18+6=24$ matches playing white, $13+2=15$ games playing
Black, and that Anand played white against Karpov a total of
$18+20+5=43$ times.  We fail to reject the null of
$w_\mathrm{An}=w_\mathrm{Kr}=w_\mathrm{Ks}$, having a likelihood ratio
of $4.5$, or an asymptotic $p$-value of $0.076$.  There is no strong
evidence that the players differ in regard to their white advantage.

@@

\cite{moul2009}: the question of Soviet collusion during the period
when the USSR dominated the struggle for the world championship has
been one of the long-standing debates in the chess world. The
best-known (but not unique) criticism was aired by Bobby Fischer some
weeks after the 1962 Candidates Tournament in Curacao

\begin{table}
\centering
\begin{tabular}{ll|cccccccc}
 &
&\rotatebox{-90}{Petrosian} 
&\rotatebox{-90}{Keres} 
&\rotatebox{-90}{Geller}
&\rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Korchnoi}
&\rotatebox{-90}{Benko}
&\rotatebox{-90}{Tal}
&\rotatebox{-90}{Filip}\\
Petrosian &USSR    &             &DDDD    &DDDD   &DWDD   &DDWW   &DDWD   &WWD    &DWWD\\
Keres     &USSR    &             &NA      &DDDD   &LDWD   &DDWD   &WWWL   &WDW    &DWWD\\
Geller    &USSR    &             &NA      &NA     &WWDL   &DDWD   &DDDW   &DWW    &DWWD\\
Fischer   &USA     &             &NA      &NA     &NA     &LWLD   &LWDW   &DWD    &WDWD\\
Korchnoi  &USSR    &             &NA      &NA     &NA     &NA     &DDDL   &WLD    &WWWW\\
Benko     &USA     &             &NA      &NA     &NA     &NA     &NA     &WLD    &LWWD\\
Tal       &USSR    &             &NA      &NA     &NA     &NA     &NA     &NA     &WLD \\
Filip     &TCH     &             &NA      &NA     &NA     &NA     &NA     &NA     &A \\
\end{tabular}
\caption{Cura\c{c}ao 1962 candidates matches, world championship 1963}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{l|cccccc}
&\rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Geller} 
&\rotatebox{-90}{Petrosian\ }
&\rotatebox{-90}{Korchnoi} 
&\rotatebox{-90}{Filip}
& $\cdots$
\\ \hline %       Fischer   Geller    Petrosian Korchnoi    Filip
Fischer    (USA)  &   -     & \halfb  & \halfw  & \oneww    &\halfb    &\\
Geller    (USSR)  & \halfw  &    -    & \halfb  & \halfw    & \halfw   &\\
Petrosian (USSR)  & \halfb  & \halfw  &    -    & \halfb    & \halfw   &\\
Korchnoi  (USSR)  & \zerob  & \halfb  & \halfw  &   -       & \oneww   &\\
Filip      (TCH)  & \halfw  & \halfb  & \halfb  & \zerob    &  -       &\\
$\vdots$&&&&&\\
\end{tabular}\label{interzonal}
\caption{Subset of the Fifth Interzonal round-robin, Stockholm 1962.
  Results (0,\halfw,1) and box colour refer to row player.  Thus, for
  example, Fischer played black against Geller and drew}
\end{table}

>>>>

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to chess-specific considerations in a number of ways.  Here I present
a new generalization of Bradley-Terry in which a chess game is
regarded as a three-way competition between the two players and an
entity that wins if the game is drawn.  Bradley-Terry can be further
modified to account for the white player advantage in a number of
ways.  Such reification affords insight into players' strengths,
response to playing black or white, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude that the
scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
evidence of collusion, in agreement with previous work.  However, I
find strong evidence of increased Soviet proclivity to draw in the
Interzonal Tournament, Stockholm 1962.
}

@@

\SweaveOpts{}
\begin{document}

@@

Bradley-Terry is not directly applicable to Chess for two reasons:
firstly, the white player is at a considerable
advantage~\cite{asdfasdffdsfds}; and secondly, many chess games end in
a draw: white wins about $37.5\%$ of all games, black wins $27.6\%$,
and about $39.4\%$ are drawn.  We consider the first-mover advantage
first, then consider the effect of drawn games.

\paragraph{First-mover advantage} Traditionally White moves first in
any game, and this feature introduces an asymmetry into play.
Historically, we see about 1.35 white wins for every black win.
However, interpretations of this statistic vary and the issue is hotly
debated with opinions ranging from Rowson who compares first-move
advantage to that of serving in tennis, to that of commentators such
as \citeauthor{adorjan} who claims that White's advantage is a
``delusion''.  \cite{asdfasdasfd} goes further, asserting that Black
can exploit the situation by mounting a directed response to White, of
the type seen in analysis of sequential zero-sum games, in which the
first-mover advantage can be positive or negative.

\paragraph{Draws} Chess games may be won by either player, or may be
drawn; the implication of a draw is that neither player is better than
the other.  Before 1867, tournament games that were drawn were
replayed and in this context we note that this procedure reduces a
chess game to a Bernoulli trial, with probability one.  This system
became unworkable due to the large number of draws, resulting in the
modern practice of awarding each player of a drawn game half a point,
rendering chess a zero-sum game.  Drawn games are generally considered
to be undesirable as they are usually poorer in spectacular or
striking elements such as sharp tactical combinations or sacrifice.
Excessive drawing dissuades sponsors, leaves spectators feeling
cheated, and gives commentators less material to discuss.  Players
themselves do not seem to value draws in the wider context of the
game: \citeauthor{fischer1969}, for example, includes only 8 draws in
his compilation of 60 memorable games, surely not representative of
tournament play in which about 39.4\% of games are drawn (binomial
test, $p<10^{-5}$).

\cite{moul2009} conceptualise a player as being able to adopt a
strategy on a spectrum ranging from conservative to aggressive, thus
being able to exert some control over the probability of a draw and
suggest that risk-averse players may pursue less vigorous strategies,
trading expected payout for reduced variance.  Further,
\cite{langer2013} point out that rational players in an ongoing
competitive situation modify their strategies depending on whether
they are leading or lagging: the lagging player will adopt a higher
risk strategy (`nothing to lose'), while the leading player is
incentivised to protect the lead by lowering the risk of losing.  In a
chess context, this means the leading player will aim for a draw.

Such arguments suggest considering a drawn chess game as a
qualitatively different outcome from a statement of equality of skill:
the play itself tends to be different, the drawing players have a
different mindset, and drawn games have less value in the wider
context of the game.

\section{Generalized Bradley-Terry}

@@

\begin{equation}
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j}\right)^{W_{ij}}
\left(\frac{p_j}{p_i+p_j}\right)^{L_{ij}}
\end{equation}

where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess game is regarded as
a three-way competition between the two players and an entity that
wins if the game is drawn.  The system is further modified to account
for first-move advantage by positing an entity whose strength is added
to that of the white player.  This methodology affords insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.

If we consider a chess game to be a three-way trial, we define a
likelihood function for non-negative $p_1,\ldots,p_n,D$ with
$D+\sum p_i=1$ and this would be

@@

The above likelihood functions are objective and reasonable
representations of chess results.  However, they differ markedly from
previous likelihood functions of draws in that they include entities
representing proclivity to draw $D$, and first-mover advantage ($W$).

\subsection{Numerical considerations}

The Method of Support \citep{edwards1972} involves only numerical
optimization (not integration which is computationally expensive).
The analysis presented here uses the framework of the \pkg{hyper2}
package~\citep{hankin2017} as a convenient and unified interface.
Because gradients are available, convergence is generally rapid and
hypothesis-specific constraints on parameter space are, in general,
straightforward to implement.

@@

Grandmasters Anand, Karpov, Kasparov have long-standing rivalry and
have played many games amongst themselves.  A summary may be found
in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &   5 &   11  \\
  Kr & 13 &  -  &    9  \\
  Ks &  2 &   7 &  -   \\
\end{tabular}
\caption{Results of 365 chess games \label{kka} between Grandmaster
  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  Thus, for
  example, Anand played white against Karpov a total of $18+20+13=51$
  times and won 18, drew 20, lost 13}
\end{table}

@@

and this is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
case: the likelihood for this hypothesis would be identically zero,
owing to the nonzero term in the numerator of the last term of
Equation~\ref{likelihood_WD}.  However, we see that the maximum
likelihood estimate for draw strength $\hat{D}\simeq 0.53$: compare
the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 games.  The MLE is lower because the
maximization process automatically adjusts for the different players'
strengths.  Secondly, we may make inferences about $W$, the extra
strength conferred by playing white: it is conceivable that $W=0$ as
this term does not appear on its own in the numerator of any term.  We
may test $H_0\colon W=0$ using the method of support by maximizing the
likelihood subject the constraint $W=0$ and comparing with the
likelihood when the optimization is free.

@@

<<>>=
ignore3
ignore4
ignore3-ignore4
@

The result of numerical optimization is about 7.01 units of support,
or a likelihood ratio of $e^{7.01}\simeq 1107.7$: that is, well over
the two-units-of-support per degree of freedom suggested by
\citeauthor{edwards1972}; we can convincingly reject the hypothesis
that $W=0$ and infer that playing white confers a significant
advantage.  This is consistent with expectation, but it is now
possible to go significantly beyond mere existence of the effect and
to quantify it, accounting for the differential strengths of the
players.  Taking the MLE above, and considering a game between say
Karpov and Anand, the strengths are Karpov=0.064, Anand=0.151, white
=0.104 and draw =0.53.  Thus we would estimate that Karpov playing
white against Kasparov would win with probability
$\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
probability $\simeq 0.174$ (compare $+0.075\ {=}0.627\ {-}0.297$ if
Karpov plays black).

@@

One very natural and interesting hypothesis would be that the three
players are of equal strength
%equal_strength_is_assessed_in_chunk_kka_equal_strength_above____________
and we might hypothesise that
$p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  This too is readily
testable by the device of maximizing the support, subject to the
constraint that $p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  In this
case the extra support in going from the constrained to the free
optimization is about 4.93, or a likelihood ratio of about $138$.
However, this time the constraint corresponds to a loss of {\em two}
degrees of freedom so we would need 4 units of support, which is
attained in this case.  Alternatively we might follow Wilks, and
observe that the value of $2\mathcal{S}=2*4.92=9.84$ lies in the tail
region of its asymptotic $\chi^2_2$ distribution with a $p$-value of
about 0.0072.  We can justifiably assert that the players do indeed
have different strengths, even if we account for white advantage and
the proclivity to draw.

@@

As argued above, players with a cautious style tend to draw and this
is an important and interesting component of the game.  To investigate
this we need a likelihood function that can account for possible
differences in draw proclivity among the players.  This may be
achieved by considering a game between two players to be a contest
between five entities: the two real players, the white advantage, and
two {\em personalised} draw entities, one for each real player; a
drawn game is counted as a win for either of the draw ``ghosts''.  Under
these assumptions, table~\ref{kka} translates into the following
likelihood function:

@@

In the likelihood function above we see, for example, that Anand won a
total of $18+6=24$ games playing white, $13+2=15$ games playing
Black, and that Anand played white against Karpov a total of
$18+20+5=43$ times.  We fail to reject the null of
$w_\mathrm{An}=w_\mathrm{Kr}=w_\mathrm{Ks}$, having a likelihood ratio
of $4.5$, or an asymptotic $p$-value of $0.076$.  There is no strong
evidence that the players differ in regard to their white advantage.

@@

The method of support allows one to account for proclivity to draw
while accounting for players' strengths and first-move advantage.  We
now consider one famous incident, the World Chess Championships 1963,
in which Fischer famously and publicly claimed that the Soviet players
colluded.  The accusation was that Soviet players agreed ahead of time
to draw when playing one another, in order to conserve energy for
subsequent matches~\citep{moul2009}.

%the question of Soviet collusion during the period when the USSR
%dominated the struggle for the world championship has been one of the
%long-standing debates in the chess world.  The best-known (but not
%unique) criticism was aired by Bobby Fischer some weeks after the 1962
%Candidates Tournament in Curacao

% \begin{table}
% \centering
% \begin{tabular}{ll|cccccccc}
% &
% &%\rotatebox{-90}{Petrosian} 
% &\rotatebox{-90}{Keres} 
% &\rotatebox{-90}{Geller}
% &\rotatebox{-90}{Fischer} 
% &\rotatebox{-90}{Korchnoi}
% &\rotatebox{-90}{Benko}
% &\rotatebox{-90}{Tal}
% &\rotatebox{-90}{Filip}\\
% Petrosian &USSR    &             &DDDD    &DDDD   &DWDD   &DDWW   &DDWD   &WWD    &DWWD\\
% Keres     &USSR    &             &NA      &DDDD   &LDWD   &DDWD   &WWWL   &WDW    &DWWD\\
% Geller    &USSR    &             &NA      &NA     &WWDL   &DDWD   &DDDW   &DWW    &DWWD\\
% Fischer   &USA     &             &NA      &NA     &NA     &LWLD   &LWDW   &DWD    &WDWD\\
% Korchnoi  &USSR    &             &NA      &NA     &NA     &NA     &DDDL   &WLD    &WWWW\\
% Benko     &USA     &             &NA      &NA     &NA     &NA     &NA     &WLD    &LWWD\\
% Tal       &USSR    &             &NA      &NA     &NA     &NA     &NA     &NA     &WLD \\
% Filip     &TCH     &             &NA      &NA     &NA     &NA     &NA     &NA     &NA \\
% \end{tabular}
% \caption{Cura\c{c}ao 1962 candidates, world championship 1963}
% \end{table}

@@

\begin{table}
\centering
\begin{tabular}{l|cccccc}
&\rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Geller} 
&\rotatebox{-90}{Petrosian\ }
&\rotatebox{-90}{Korchnoi} 
&\rotatebox{-90}{Filip}
& $\cdots$
\\ \hline %       Fischer   Geller    Petrosian Korchnoi    Filip
Fischer    (USA)  &   -     & \halfb  & \halfw  & \oneww    &\halfb    &\\
Geller    (USSR)  & \halfw  &    -    & \halfb  & \halfw    & \halfw   &\\
Petrosian (USSR)  & \halfb  & \halfw  &    -    & \halfb    & \halfw   &\\
Korchnoi  (USSR)  & \zerob  & \halfb  & \halfw  &   -       & \oneww   &\\
Filip      (TCH)  & \halfw  & \halfb  & \halfb  & \zerob    &  -       &\\
$\vdots$&&&&&\\
\end{tabular}\label{interzonal}
\caption{Subset of the Fifth Interzonal round-robin, Stockholm 1962.
  Results $\in\{0,\halfw,1\}$ and box colour refer to row
  player.  Thus, for example, Fischer played black against Geller and drew}
\end{table}
